\section{Top quark and W boson}
  \subsection{CKM and W-boson}
	The CKM-matrix is a unitary matrix where each element holds information about the
      	strength of the flavour changing weak deacys which happens between quarks.
      	These changes are mediated with the $W^{\pm}$ boson.
      	When four quarks were discovered it was created two sets of equation describing
      	the decay from down and strange into top and charm.
      	Seeing that with CP-violation could not be explained with these four quarks,
      	they added another generation to create the CKM-matrix:
      	\begin{equation}
      	\begin{bmatrix}
      	  d'\\
      	  s'\\
      	  b'
      	\end{bmatrix}
      	=
      	\begin{bmatrix}
      	  V_{ud} & V_{us} & V_{ub} \\
      	  V_{cd} & V_{cs} & V_{cb} \\
      	  V_{td} & V_{ts} & V_{tb} 
      	\end{bmatrix}
      	\begin{bmatrix}
      	  d\\
      	  s\\
      	  b
      	\end{bmatrix}
      	\end{equation}
      	The $W$ boson is a mediater of the weak force, it has either $+1$ or $-1$ charge so it
	can react with charged particles. In regards to the CKM-matrix it is the mediator for
	decaying quarks between up and down types as well as changing flavours.
	$V_{ud}$ can be experimentally shown from the ration between netron decay and
	$\mu$ decay.\\
	$V_{us}$ is shown in the $K^+ \rightarrow \pi^0 e^+ \nu_e$ decay process.\\
	$V_{cs}$ is experimentally shown in hadronic decays of $W^{\pm}$ and 
	$D\rightarrow \bar{K}e^+\nu_e$ process.
  \subsection{Top quark production}

	In an $e^-e_+$ anihilation there will be produced a $\gamma$ or $Z$
	which in turn can decay into a $q\bar{q}$. This pair can be a $t\bar{t}$ but is
	not favorable due to its high mass.

	As for the $pp$-collison there are alot that can happen. It can produce $e^-e^+$ which was 
	discussd above. Gluons, $g$, can be produced which in turn decay into $t\bar{t}$-pair,
	and $g$ going into $b\bar{b}$ in which one of the $b$'s interacts with a $W$ to become
	a singel $t$.

	When a $p\bar{p}$ annihilation occurs there are high probability of high energy $g$ to
	be produced which have enough energy to decay into $t\bar{t}$, and $b\bar{b}$ to $W$ and $bt$
	for a single top.

  \subsection{Top quark decay}

    The top quark is very heavy compared to the other elementary particles in the standard model.
    Due to this mass it only as a lifetime of $5\times 10^{-25}\,s$, and therefore do not have time
    to create hadrons, and is the only quark we can observe alone in some sense. 
    Because of that it deacys into a $W$ and a bottom, strange or down quark, which is also the only 
    observed decay mode of the $t$ quark.

    The large mass sets the $t$ apart from the other quarks, it decays too fast to hadronize, 


  \subsection{Branching ratios of top decays}
    \[ W_i = \frac{\Gamma_i/\Gamma}{\tau} \approx G_F (\Delta m)^5\]
    $G_F \approx 1.17 \times 10^{-5}\,GeV^{-2}$\\
    The branching ratio will be about the size of the mass difference in the process.
    \[m_t \approx 173\,GeV,\,m_b \approx 4\,GeV,\, m_c \approx 1\,GeV,\, m_s \approx %
      95\,MeV,\, m_{\tau} \approx 1776\,MeV \approx 2\,GeV\]
    $t\rightarrow b+c\bar{s}$
    \[ \Delta m = m_t - m_b - m_c - m_s  \approx  168\,GeV\]
    $t\rightarrow b + \tau^+\nu_{\tau}$ aproximate the neutrino to be massless.
    \[ \Delta m = m_t - m_b - m_{\tau^+}  \approx  167\,GeV\]
    
