\section{Gauge theories}
  \subsection{Symmetries}

    The Standard Model is based on the symmetry group $SU(3)\otimes %
    SU(2) \otimes U(1)$.
    \begin{itemize}
      \item
	$QED$ is described in the $U(1)$ group giving rise to the photon as a mediating gauge boson.

      \item
	$WI$ from $SU(2)$ which produces three bosons, $W^{\pm},\,W^0$, the $Z$ will come after
    	combining $SU(2)_L\otimes U(1)_Y$ into electro weak theory. In electro weak theory it can
	be shown conservation of isospin.  

      \item
	And finally the $QCD$ comes from the $SU(3)_C$ where the 8 bosons are gluons,
    	and it holds the conservation of the quantum number color. 
    \end{itemize} 
    
    The fermions are organized in three families as seen below.
    \[%
        \begin{bmatrix}
          \nu_e & u \\
          e^- & d'
        \end{bmatrix}%
        ,\quad%
        \begin{bmatrix}
          \nu_{\mu} & c\\
          \mu^- & s'
        \end{bmatrix}%
        ,\quad%
        \begin{bmatrix}
          \nu_{\tau} & t\\
          \tau^- & b'
        \end{bmatrix}%
    \]
    And each family can be divided into left handed doublets and right handed
    singlets $SU(2)_L$. There is the foce mediators $g$, $W$, $Z$, $\gamma$ and 
    the $H$ boson which gives particles mass.
    \paragraph{GUT}
    In $GUT$ the coupling constants will all be the same, meaning that all the forces will
    have equal strenght in their coupling. These are the conditions are considered to be
    available at very high energies, as in the start of our universe.

    \subsection{Gauge principle and $QCD$}
      
      The gauge principle is a requirement that the phase invariance should hold locally.
      These transformations give rise to mediator bosons.
      
      Starting with the free Lagrangian
      \begin{equation}
	\mathcal{L}_0 = \sum_f \bar{q}_f(i\gamma^{\mu}\partial_{\mu} - m_f)q_f
      \end{equation}
      where $q^{\alpha}_f \equiv (q^1_f,q^2_f,q^3_f)$ has colour $\alpha$ and flavour $f$. 
      This Lagrangian is invariant under global transformations in $SU(3)_C$ in colour space,
      \begin{equation}
	q^{\alpha}_f \rightarrow U^{\alpha}_{\beta}q^{\beta}_f,\quad%
	U= \exp\left\{i\frac{\lambda^a}{2}\theta_a\right\}
      \end{equation}
      here $U$ are the $SU(3)_C$ matrices, $\lambda$ are the generators in the fundemental
      representation.

      Now we require the Lagrangian to be also invariant under local $SU(3)_C$ transformations,
      which is done in the same way as $QED$, by changing the deriviatives to covariant objects.
      \begin{equation}
	D^{\mu}q_f \equiv \big[ \partial^{\mu} + ig_s G^{\mu}(x)\big]q_f
      \end{equation}
      Here the $G^{\mu}_a$ are the eight different bosons in $QCD$, also called gluons.

      Since we want the transformation to be exactly like the transform of the colour-vector $q_f$,
      we get
      \begin{flalign}
	&(D^{\mu})' \rightarrow UD^{\mu}U^{\dagger}\\
	&(G^{\mu}_a)' \rightarrow UG^{\mu}U^{\dagger} + \frac{i}{g_s}(\partial^{\mu}U)U^{\dagger}
      \end{flalign}
	
      Then under an infinitesimal $SU(3)_C$ transformation of the quarks and bosons
      \begin{flalign}
	&U \approx 1 + i\left(\frac{\lambda^a}{2}\right)_{\alpha\beta}\\
	&(q^{\alpha}_f)' \rightarrow q^{\alpha}_f +%
	i\left(\frac{\lambda^a}{2}\right)_{\alpha\beta}\delta\theta_aq^{\beta}_f\\
	&(G^{\mu}_a)' \rightarrow \frac{1}{g_s}\partial^{\mu}(\delta\theta_a)%
	- f^{abc}\delta\theta)_bG^{\mu}_c
      \end{flalign}

      We need to create field strengths of the gluon fields who transform homogeneously 
      \[ (G^{\mu\nu})' = UG^{\mu\nu}U^{\dagger}\]
      
      \begin{flalign}
	&G^{\mu\nu}(x) \equiv \frac{\lambda^a}{2}G^{\mu\nu}_a(x)\\
	&G^{\mu\nu}_a(x) = \partial^{\mu}G^{\nu}_a - \partial^{\nu}_a%
	- g_s f^{abc}G^{\mu}_b G^{\nu}_c
      \end{flalign}
      and taking the trace of the kinetic term og the gluon fields %
      $\operatorname{Tr}[G^{\mu\nu}G_{\mu\nu}] = \frac{1}{2} G^{\mu\nu}_a G_{\mu\nu}^a$.
      All this results in the Lagrangian of $QCD$
      \begin{equation}
	\mathcal{L}_{QCD} \equiv -\frac{1}{4}G^{\mu\nu}_a G^a_{\mu\nu} + %
	\sum_f \bar{q}_f(i\gamma^{\mu}D_{\mu} - m_f)q_f
      \end{equation}
      
      As with $QED$ there is no invariant mass term for the gluon fields, making them massless
      spin-1 bosons. There is also self interacting terms due to the non-commutativity of the
      $SU(3)_C$ matrices, which do not occur in $QED$.

      

